\input{../../resources/lesson-head.tex}

\section*{Let's get started}

We want you to be successful. 

We will work together to build an 
environment in CSE 105 that supports your learning
in a way that respects your
perspectives, experiences, and identities (including race, ethnicity, heritage, gender, sex, 
class, sexuality, religion, ability, age, educational background, etc.).  
Our goal is for you to  engage
with interesting and challenging concepts and 
feel comfortable exploring, asking questions, and thriving.\\

If you are skipping and stretching meals, or having difficulties affording or accessing food, you may be eligible for CalFresh, California's Supplemental Nutrition Assistance Program, that can provide up to \$292 a month in free money on a debit card to buy food. Students can apply at \url{benefitscal.com/r/ucsandiegocalfresh}.
The Hub Basic Needs Center empowers all students by connecting them to resources for food, stable housing and financial literacy. Visit their site at \url{basicneeds.ucsd.edu}.

Financial aid resources, the possibility of emergency grant funding, and off-campus housing referral 
resources are available: see your College Dean of Student Affairs.

If you find yourself in an uncomfortable situation, ask for help. 
We are committed to upholding University policies regarding nondiscrimination, sexual violence and sexual harassment.
Here are some campus contacts that could provide this help: 
Counseling and Psychological Services (CAPS) at 858 534-3755 or \href{http://caps.ucsd.edu}{http://caps.ucsd.edu}; 
OPHD at 858 534-8298 or ophd@ucsd.edu , \href{http://ophd.ucsd.edu}{http://ophd.ucsd.edu};
CARE at Sexual Assault Resource Center at 858 534-5793 or sarc@ucsd.edu , \href{http://care.ucsd.edu}{http://care.ucsd.edu}.


Please reach out (minnes@ucsd.edu) if you need support with extenuating circumstances affecting CSE 105.

\vfill

\section*{Introductions}
Class website on Canvas \href{https://canvas.ucsd.edu/}{https://canvas.ucsd.edu}


Instructor: Prof. Mia Minnes {\tiny{"Minnes" rhymes with Guinness}}, minnes@ucsd.edu, 
\href{http://cseweb.ucsd.edu/~minnes}{http://cseweb.ucsd.edu/~minnes}


Our team: One instructor + two TAs and five tutors + all of you

Fill in contact info for students around you, if you'd like:


\vfill

\newpage 
Welcome to CSE 105: Introduction to Theory of Computation in Winter 2025!

\section*{CSE 105's Big Questions}
\begin{itemize}
   \item What problems are computers capable of solving?
   \item What resources are needed to solve a problem?
   \item Are some problems harder than others?
\end{itemize}

In this context, a {\bf problem} is defined as: ``Making a decision or computing a value based on some input"

Consider the following problems: 
\begin{itemize}
   \item Find a file on your computer
   \item Determine if your code will compile
   \item Find a run-time error in your code
   \item Certify that your system is un-hackable
\end{itemize}

Which of these is hardest?

\vfill

In Computer Science, we operationalize ``hardest'' as ``requires most resources'', where
resources might be memory, time, parallelism, randomness, power, etc.

To be able to compare ``hardness'' of problems, we use a consistent description of problems

{\bf Input}: String

{\bf Output}: Yes/ No, where Yes means that the input string matches the pattern or property described by the problem.

\newpage

\subsection*{Weeks 0 and 1 at a glance}

\subsubsection*{Textbook reading: Chapter 0, Sections 1.3, 1.1}

\vspace{-15pt}

Before Monday, review class syllabus on Canvas (https://canvas.ucsd.edu/).

Before Wednesday, read Example 1.51.

{\it Notice}: we are jumping to Section 1.3 and then will come back 
to Section 1.1 next week.

Before Friday, read Definition 1.52 (definition of regular expressions) on page 64.

For Week 2 Monday: Figure 1.4 and Definition 1.5 (definition of finite automata) on pages 34-35.

{\it Textbook references: Within a chapter, each item is numbered consecutively. Example 1.51 is the fifty-first numbered item in chapter one.}


\subsubsection*{We will be learning and practicing to:}
%Define decision problem, Regular expressions
\begin{itemize}
\item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
\begin{itemize}
   \item Translate a decision problem to a set of strings coding the problem.
   \begin{itemize}
      \item {\bf Distinguish between alphabet, language, sets, and strings}
   \end{itemize}
   \item Use regular expressions and relate them to languages and automata.
   \begin{itemize}
      \item {\bf Write and debug regular expressions using correct syntax}
      \item {\bf Determine if a given string is in the language described by a regular expression}
   \end{itemize}
\end{itemize}
\end{itemize}

\subsubsection*{TODO:}
\begin{list}{\itemsep-10pt}
   \item \#FinAid Assignment on Canvas (complete as soon as possible) and read syllabus on Canvas
   \item Schedule your Test 1 Attempt 1, Test 2 Attempt 1, Test 1 Attempt 2, and Test 2 Attempt 2 times 
   at PrairieTest (http://us.prairietest.com)
   \item Review Quiz 1 on PrairieLearn (http://us.prairielearn.com), complete by 1/15/25
   \item Create a homework group, possibly by using the Piazza (https://piazza.com/) find-a-teammate tool
   \item Homework 1 submitted via Gradescope (https://www.gradescope.com/), due 1/16/25
\end{list}

\newpage

\section*{Week 1 Monday: Terminology and Notation}

\input{../activity-snippets/day1.tex}
    
\newpage

\section*{Week 1 Wednesday: Regular expressions}

\input{../activity-snippets/day2.tex}
    
\newpage

\subsection*{Week 1 Friday: Regular expressions conventions}

\input{../activity-snippets/day3.tex}



\newpage

\subsection*{Week 2 at a glance}

\vspace{-15pt}

\subsubsection*{Textbook reading: Sections 1.1, 1.2}

\vspace{-15pt}

Before Monday, read Figure 1.4 and Definition 1.5 (definition of finite automata) on pages 34-35.

Before Wednesday, read pages 41-43 (Figures 1.18, 1.19, 1.20) for examples of automata and languages. Wednesday's class will be asynchronous. Please read over the annotated notes and watch the relevant supplementary videos. Ask any followup questions in the discussion forum or in office hours.

Before Friday, read pages 48-50 (Figures 1.27, 1.29) which introduces nondeterminism.

No class on Week 3 Monday in observance of Martin Luther King Jr. Day.

For Week 3 Wednesday: read the definition of the union, concatenation, and star operations for languages,  given 
as Definition 1.23 on page 44 and a useful example is Example 1.24.

\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}
%Define decision problem, Formal definition of automata, Informal definition of automata, Find example languages, 
%Describe and use models of computation that don't involve state machines.
\begin{itemize}
\item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
\begin{itemize}
   \item Give examples of sets that are regular (and prove that they are).
   \begin{itemize}
      \item {\bf State the definition of the class of regular languages}
      \item {\bf Give examples of regular languages, using each of the three equivalent models of computation for proving regularity.}
   \end{itemize}
   \item Describe and use models of computation that don't involve state machines.
   \begin{itemize}
      \item {\bf Given a DFA, find a regular expression that describes its language.}
      \item {\bf Given a regular expression, find a DFA that recognizes its language.}
   \end{itemize}
   \item Use precise notation to formally define the state diagram of finite automata.
   \item Use clear English to describe computations of finite automata informally.
      \begin{itemize}
         \item {\bf State the formal definition of (deterministic) finite automata}   
         \item {\bf Trace the computation of a finite automaton on a given string using its state diagram}
         \item {\bf Translate between a state diagram and a formal definition}
         \item {\bf Determine if a given string is in the language recognized by a finite automaton}
         \item {\bf Design an automaton that recognizes a given language}
         \item {\bf Specify a general construction for DFA based on parameters}
         \item {\bf Design general constructions for DFA}
   \end{itemize}
\end{itemize}

\end{itemize}
\newpage
\subsubsection*{TODO:}
\begin{list}{\itemsep-10pt}
   \item \#FinAid Assignment on Canvas (complete as soon as possible) and read syllabus on Canvas
   \item Schedule your Test 1 Attempt 1, Test 2 Attempt 1, Test 1 Attempt 2, and Test 2 Attempt 2 times 
   at PrairieTest (http://us.prairietest.com)
   \item Review Quiz 1 on PrairieLearn (http://us.prairielearn.com), complete by 1/15/25
   \item Create a homework group, possibly by using the Piazza (https://piazza.com/) find-a-teammate tool
   \item Homework 1 submitted via Gradescope (https://www.gradescope.com/), due 1/16/25
   \item Review Quiz 2 on PrairieLearn (http://us.prairielearn.com), complete by 1/22/25
\end{list}


\newpage

\subsection*{Week 2 Monday: Finite automata}

\input{../activity-snippets/day4.tex}


\subsection*{Week 2 Wednesay: Finite automaton constructions - asynchronous}

\input{../activity-snippets/day5.tex}

\newpage
\subsection*{Week 2 Friday: Nondeterministic automata}

We saw that whenever a language is recognized by a DFA, its
complement is also recognized by some (other) DFA. 

Another way to say this is that the collection of languages
that are each recognizable by a DFA is {\bf closed} under complementation.


Before we continue with more complicated constructions, let's revisit some of 
the assumption of the {\bf D}FA model.

Recall that the computation of {\bf deterministic} finite automaton has exactly one choice for its next step given the current state and the current character to read.


\input{../activity-snippets/day6.tex}



\newpage

\subsection*{Week 3 at a glance}


\subsubsection*{Textbook reading: Chapter 1}

\vspace{-15pt}

No class on Week 3 Monday in observance of Martin Luther King Jr. Day.

Before Wednesday: read the definition of the union, concatenation, and star operations for languages,  given 
as Definition 1.23 on page 44 and a useful example is Example 1.24.

Before Friday,  read pages 45-46 (Theorem 1.25) that we'll refer to as a ``closure proof".

For Week 4 Monday, read Introduction to Section 1.4 (page 77) which introduces nonregularity.


\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}

\vspace{-15pt}

%Define decision problem, Formal definition of automata, Informal definition of automata, Find example languages, 
%Describe and use models of computation that don't involve state machines.
\begin{itemize}
\item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
\begin{itemize}
   \item Use precise notation to formally define the state diagram of finite automata.
   \item Use clear English to describe computations of finite automata informally.
      \begin{itemize}
         \item {\bf Motivate the use of nondeterminism}
         \item {\bf State the formal definition of NFA}   
         \item {\bf Trace the computation(s) of a NFA on a given string using its state diagram}
         \item {\bf Determine if a given string is in the language recognized by a NFA}
         \item {\bf Translate between a state diagram and a formal definition of a NFA}
      \end{itemize}
   \item Give examples of sets that are regular (and prove that they are).
   \begin{itemize}
      \item {\bf State the definition of the class of regular languages}
      \item {\bf Give examples of regular languages, using each of the three equivalent models of computation for proving regularity.}
      \item {\bf Choose between multiple models to prove that a language is regular}
      \item {\bf Explain the limits of the class of regular languages}
   \end{itemize}
   \item Describe and use models of computation that don't involve state machines.
   \begin{itemize}
      \item {\bf Given a DFA or NFA, find a regular expression that describes its language.}
      \item {\bf Given a regular expression, find a DFA or NFA that recognizes its language.}
   \end{itemize}
\end{itemize}


\item Understand, guide, shape impact of computing on society/the world. Connect the role of Theory CS classes to other applications (in undergraduate CS curriculum and beyond). Model problems using appropriate mathematical concepts.
 \begin{itemize}
     \item Explain nondeterminism and describe tools for simulating it with deterministic computation.
     \begin{itemize}
       \item {\bf Given a NFA, find a DFA that recognizes its language.}
       \item {\bf Convert between regular expressions and automata}
     \end{itemize}
 \end{itemize}

\end{itemize}

\vspace{-20pt}

\subsubsection*{TODO:}
\begin{list}{\itemsep-10pt}
   \item Schedule your Test 1 Attempt 1, Test 2 Attempt 1, Test 1 Attempt 2, and Test 2 Attempt 2 times 
   at PrairieTest (http://us.prairietest.com)
   \item Review Quiz 3 on PrairieLearn (http://us.prairielearn.com), due 1/29/2025
   \item Homework 2 submitted via Gradescope (https://www.gradescope.com/), due Tuesday 1/30/2025
\end{list}


\vfill

In Computer Science, we operationalize ``hardest'' as ``requires most resources'', where
resources might be memory, time, parallelism, randomness, power, etc.
To be able to compare ``hardness'' of problems, we use a consistent description of problems

{\bf Input}: String

{\bf Output}: Yes/ No, where Yes means that the input string matches the pattern or property described by the problem.

So far: we saw that regular expressions are convenient ways of describring patterns in strings.
{\bf Finite automata} give a model of computation for processing strings and and classifying them into Yes (accepted)
or No (rejected). We will see that each set of strings is described by a regular expression if and only 
if there is a FA that recognizes it.  Another way of thinking about it: properties described by regular
expressions require exactly the computational power of these finite automata.

\newpage
\subsection*{Wednesday: Automata constructions}

\input{../activity-snippets/day7.tex}

\newpage
\subsection*{Friday: Regular langauges}

\input{../activity-snippets/day8.tex}

\newpage

\subsection*{Week 4 at a glance}

\subsubsection*{Textbook reading: Section 1.4, 2.2, 2.1.}

\vspace{-15pt}

Before Monday, read Introduction to Section 1.4 (page 77) which introduces nonregularity.

Before Wednesday, read Definition 2.13 (page 111-112) introducing Pushdown Automata.

Before Friday, read Example 2.18 (page 114).

For Week 5 Monday: read Introduction to Section 2.1 (pages 101-102).

\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}

\begin{itemize}
    \item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
    \begin{itemize}
       \item Give examples of sets that are regular (and prove that they are).
       \begin{itemize}
          \item {\bf State the definition of the class of regular languages}
          \item {\bf Explain the limits of the class of regular languages}
          \item {\bf Identify some regular sets and some nonregular sets}
       \end{itemize}
       \item Use precise notation to formally define the state diagram of a PDA
       \item Use clear English to describe computations of PDA informally.
       \begin{itemize}
           \item {\bf Define push-down automata informally and formally}
           \item {\bf State the formal definition of a PDA}
           \item {\bf Trace the computation(s) of a PDA on a given string using its state diagram}
           \item {\bf Determine if a given string is in the language recognized by a PDA}
           \item {\bf Translate between a state diagram and a formal deﬁnition of a PDA}
           \item {\bf Determine the language recognized by a given PDA}
        \end{itemize}

    \end{itemize}

    \item Know, select and apply appropriate computing knowledge and problem-solving techniques. 
    \begin{itemize}
    \item Apply classical techniques including pumping lemma, determinization, diagonalization, and reduction to analyze the complexity of languages and problems.
    \begin{itemize}
        \item {\bf Justify why the Pumping Lemma is true.}
        \item {\bf Use the pumping lemma to prove that a given language is not regular.}
    \end{itemize}
    \end{itemize}
\end{itemize}

\vspace{-20pt}

\subsubsection*{TODO:}
\begin{list}{\itemsep-10pt}
   \item Schedule your Test 1 Attempt 1, Test 2 Attempt 1, Test 1 Attempt 2, and Test 2 Attempt 2 times 
   at PrairieTest (http://us.prairietest.com)
   \item Review Quiz 3 on PrairieLearn (http://us.prairielearn.com), due 1/29/2025
   \item Homework 2 submitted via Gradescope (https://www.gradescope.com/), due 1/30/2025
   \item Review Quiz 4 on PrairieLearn (http://us.prairielearn.com), due 2/5/2025
\end{list}

\newpage
\subsection*{Monday: Pumping Lemma}

\input{../activity-snippets/day9.tex}

\newpage
\subsection*{Wednesday: Proving nonregularity, and beyond}

\input{../activity-snippets/day10.tex}
    
\newpage
\subsection*{Friday: Pushdown Automata}

\input{../activity-snippets/day11.tex}


\newpage

\subsection*{Week 5 at a glance}

\subsubsection*{Textbook reading: Chapter 2}

\vspace{-20pt}

Before Monday, read Introduction to Section 2.1 (pages 101-102).

Before Wednesday, read Section 2.1

Before Friday, read Theorem 2.20.

For Week 6 Monday: Page 165-166 Introduction to Section 3.1.

\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}
\vspace{-20pt}

\begin{itemize}
    \item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
    \begin{itemize}
        \item Describe and use models of computation that don't involve state machines.
        \begin{itemize}
            \item {\bf Identify the components of a formal definition of a context-free grammar (CFG)}
            \item {\bf Derive strings in the language of a given CFG}
            \item {\bf Determine the language of a given CFG}
            \item {\bf Design a CFG generating a given language}
            \item {\bf Use context-free grammars and relate them to languages and pushdown automata.}
        \end{itemize}
        \item Use precise notation to formally define the state diagram of a Turing machine
        \item Use clear English to describe computations of Turing machines informally.
        \begin{itemize}
            \item {\bf Design a PDA that recognizes a given language.}
         \end{itemize}
       \item Give examples of sets that are context-free (and prove that they are).
       \begin{itemize}
          \item {\bf State the definition of the class of context-free languages}
          \item {\bf Explain the limits of the class of context-free languages}
          \item {\bf Identify some context-free sets and some non-context-free sets}
       \end{itemize}
    \end{itemize}
    \item Know, select and apply appropriate computing knowledge and problem-solving techniques. 
    Reason about computation and systems.
    \begin{itemize}
        \item Describe and prove closure properties of classes of languages under certain operations.
        \begin{itemize}
            \item {\bf Apply a general construction to create a new PDA or CFG from an example one.}
            \item {\bf Formalize a general construction from an informal description of it.}
            \item {\bf Use general constructions to prove closure properties of the class of context-free langugages.}
            \item {\bf Use counterexamples to prove non-closure properties of the class of context-free langugages.}
        \end{itemize}
    \end{itemize}
\end{itemize}

\vspace{-20pt}

\subsubsection*{TODO:}
\begin{list}{\itemsep-10pt}
   \item Schedule your Test 1 Attempt 1, Test 2 Attempt 1, Test 1 Attempt 2, and Test 2 Attempt 2 times
   at PrairieTest (http://us.prairietest.com) . The first Test 1 sessions are next week!
   \item Review Quiz 4 on PrairieLearn (http://us.prairielearn.com), due 2/5/2025
   \item Homework 3 submitted via Gradescope (https://www.gradescope.com/), due 2/6/2025
   \item Review Quiz 5 on PrairieLearn (http://us.prairielearn.com), due 2/12/2025
\end{list}

\newpage



\newpage
\subsection*{Monday: More Pushdown Automata}

\input{../activity-snippets/day12.tex}

\newpage
\section*{Wednesday: Context-free Grammars and Languages}

\input{../activity-snippets/day13.tex}
    
\newpage
\subsection*{Friday: Context-free and non-context-free languages}

\input{../activity-snippets/day14.tex}




\newpage

\subsection*{Week 6 at a glance}

\vspace{-10pt}

\subsubsection*{Textbook reading: Chapter 3}

\vspace{-10pt}

Before Monday, Page 165-166 Introduction to Section 3.1.

\vspace{-10pt}

Before Wednesday, Example 3.9 on page 173.

\vspace{-10pt}

Before Friday, Page 184-185 Terminology for describing Turing machines.

\vspace{-10pt}

For Week 7 Monday: Introduction to Chapter 4.

\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}
\vspace{-20pt}

\begin{itemize}
    \item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
    \begin{itemize}
        \item Use precise notation to formally define the state diagram of a Turing machine
        \item Use clear English to describe computations of Turing machines informally.
        \begin{itemize}
                \item {\bf Motivate the definition of a Turing machine}
                \item {\bf Trace the computation of a Turing machine on given input}
                \item {\bf Describe the language recognized by a Turing machine}
                \item {\bf Determine if a Turing machine is a decider}
                \item {\bf Given an implementation-level description of a Turing machine}
                \item {\bf Use high-level descriptions to define and trace Turing machines}
                \item {\bf Apply dovetailing in high-level definitions of machines}
         \end{itemize}
       \item Give examples of sets that are recognizable and decidable (and prove that they are).
       \begin{itemize}
          \item {\bf State the definition of the class of recognizable languages}
          \item {\bf State the definition of the class of decidable languages}
          \item {\bf State the definition of the class of co-recognizable languages}
       \end{itemize}
    \end{itemize}
    \item Know, select and apply appropriate computing knowledge and problem-solving techniques. 
    Reason about computation and systems.
    \item Describe and prove closure properties of classes of languages under certain operations.
    \begin{itemize}
        \item {\bf Apply a general construction to create a new Turing machine from an example one.}
        \item {\bf Formalize a general construction from an informal description of it.}
        \item {\bf Use general constructions to prove closure properties of the class of decidable langugages.}
        \item {\bf Use general constructions to prove closure properties of the class of recognizable langugages.}
    \end{itemize}
\end{itemize}

\vspace{-20pt}

\subsubsection*{TODO:}
\begin{list}{\itemsep -20pt}
   \item This week: Test 1 Attempt 1 in CBTF. 
   \item Review Quiz 5 on PrairieLearn (http://us.prairielearn.com), due 2/12/2025
   \item Mid quarter feedback survey on Canvas.
\end{list}

\newpage

\section*{Monday: Descriptions of Turing machines}

\input{../activity-snippets/day15.tex}
    
\newpage
\subsection*{Wednesday: Recognizable and decidable languages}

\input{../activity-snippets/day16.tex}


\newpage
\subsection*{Friday: Closure for the classes of recognizable and decidable languages}

\input{../activity-snippets/day17.tex}



\newpage

\subsection*{Week 7 at a glance}

\subsubsection*{Textbook reading: Chapter 4}

\vspace{-20pt}
No class on Monday in observance of UCSD holiday.

Before Wednesday, Introduction to Chapter 4.

Before Friday, Decidable problems concerning regular languages, 
Sipser pages 194-196.

For Week 8 Monday: An undecidable language, Sipser pages 207-209.

\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}

\vspace{-20pt}

\begin{itemize}
    \item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
    \begin{itemize}
        \item Use clear English to describe computations of Turing machines informally.
        \begin{itemize}
                \item {\bf Use high-level descriptions to define and trace Turing machines}
                \item {\bf Apply dovetailing in high-level definitions of machines}
        \end{itemize}
        \item Give examples of sets that are regular, context-free, decidable, or recognizable (and prove that they are).
        \begin{itemize}
          \item {\bf Give examples of sets that are decidable.}
          \item {\bf Give examples of sets that are recognizable.}
       \end{itemize}
    \end{itemize}
    \item Know, select and apply appropriate computing knowledge and problem-solving techniques. Reason about computation and systems.
    \begin{itemize}
        \item Translate a decision problem to a set of strings coding the problem.
        \begin{itemize}
        \item {\bf Connect languages and computational problems}
        \item {\bf Describe and use the encoding of objects as inputs to Turing machines}
        \item {\bf Trace high-level descriptions of algorithms for computational problems}
        \end{itemize}
    \item Classify the computational complexity of a set of strings by determining whether it is regular, context-free, decidable, or recognizable.
    \begin{itemize}
    \item {\bf Describe common computational problems with respect to DFA, NFA, regular expressions, PDA, and context-free grammars.}
    \item {\bf Give high-level descriptions of Turing machines that decide common computational problems with respect to DFA, NFA, regular expressions, PDA, and context-free grammars.}
\end{itemize}
\end{itemize}
\end{itemize}

\vspace{-20pt}

\subsubsection*{TODO:}
\begin{list}{\itemsep-10pt}
    \item Review Quiz 6 on PrairieLearn (http://us.prairielearn.com), due 2/19/2025
    \item Homework 4 submitted via Gradescope (https://www.gradescope.com/), due 2/20/2025
    \item Review Quiz 7 on PrairieLearn (http://us.prairielearn.com), due 2/26/2025
\end{list}

\newpage

\subsection*{Monday: No class, in observance of UCSD holiday}
\subsection*{Wednesday: General constructions for Turing machines}


\input{../activity-snippets/day18.tex}


\newpage
\subsection*{Friday: Decidable problems about regular languages}

\input{../activity-snippets/day19.tex}



\newpage

\subsection*{Week 8 at a glance}


\vspace{-20pt}

\subsubsection*{Textbook reading: Chapter 4, Section 5.3}

\vspace{-20pt}

Before Monday, ``An undecidable language", Sipser pages 207-209.

Before Wednesday, Definition 5.20 and figure 5.21 (page 236) of mapping reduction.

Before Friday, Example 5.24 (page 236).

For Week 9 Monday: Example 5.26 (page 237).

\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}

\vspace{-20pt}

\begin{itemize}
    \item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
    \begin{itemize}
        \item Give examples of sets that are regular, context-free, decidable, or recognizable (and prove that they are).
        \begin{itemize}
          \item {\bf Define and explain the definitions of the computational problem $A_{TM}$}
          \item {\bf Define and explain the definitions of the computational problem $HALT_{TM}$}
       \end{itemize}
    \end{itemize}
    \item Know, select and apply appropriate computing knowledge and problem-solving techniques. Reason about computation and systems.
    \begin{itemize}
        \item Use diagonalization to prove that there are 'hard' languages relative to certain models of computation.
        \begin{itemize}
            \item {\bf Trace the argument that proves $A_{TM}$ is undecidable and explain why it works.}
        \end{itemize}
        \item Use mapping reduction to deduce the complexity of a language by comparing to the complexity of another.
           \begin{itemize}
              \item {\bf Define computable functions, and use them to give mapping reductions between computational problems}
              \item {\bf Build and analyze mapping reductions between computational problems}
              \item {\bf Deduce the decidability or undecidability of a computational problem given mapping reductions between it and other computational problems, or explain when this is not possible.}
           \end{itemize}
    \item Classify the computational complexity of a set of strings by determining whether it is regular, context-free, decidable, or recognizable.
    \begin{itemize}
    \item {\bf State, prove, and use theorems relating decidability, recognizability, and co-recognizability.}
    \item {\bf Prove that a language is decidable or recognizable by defining and analyzing a Turing machines with appropriate properties.}
\end{itemize}
\end{itemize}
\end{itemize}

\vspace{-20pt}

\subsubsection*{TODO:}
\begin{list}{\itemsep-10pt}
   \item Review Quiz 7 on PrairieLearn (http://us.prairielearn.com), due 2/26/2025
   \item Homework 5 submitted via Gradescope (https://www.gradescope.com/), due 2/27/2025
   \item Review Quiz 8 on PrairieLearn (http://us.prairielearn.com), due 3/5/2025
\end{list}

\newpage

\section*{Monday: $A_{TM}$ is recognizable but undecidable}

\input{../activity-snippets/day20.tex}
    
\newpage
\subsection*{Wednesday: Computable functions and mapping reduction}

\input{../activity-snippets/day21.tex}


\newpage
\subsection*{Friday: The Halting problem}

\input{../activity-snippets/day22.tex}

\newpage

\subsection*{Week 9 at a glance}

\vspace{-10pt}

\subsubsection*{Textbook reading: Section 5.3, Section 5.1, Section 3.2}

\vspace{-10pt}

For Monday, Example 5.26 (page 237).

For Wednesday,  Theorem 5.30 (page 238) 

For Friday, skim section 3.2.

{\it For Monday of Week 10}: Definition 7.1 (page 276)
\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}

\vspace{-20pt}

\begin{itemize}
    \item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
    \begin{itemize}
        \item Give examples of sets that are regular, context-free, decidable, or recognizable (and prove that they are).
        \begin{itemize}
          \item {\bf Define and explain computational problems, including $A_{**}$, $E_{**}$, $EQ_{**}$, (for **  DFA or TM) and $HALT_{TM}$}
       \end{itemize}
    \end{itemize}
    \item Know, select and apply appropriate computing knowledge and problem-solving techniques. Reason about computation and systems.
    \begin{itemize}
        \item Use mapping reduction to deduce the complexity of a language by comparing to the complexity of another.
           \begin{itemize}
              \item {\bf Explain what it means for one problem to reduce to another}
              \item {\bf Define computable functions, and use them to give mapping reductions between computational problems}
              \item {\bf Build and analyze mapping reductions between computational problems}
           \end{itemize}
    \item Classify the computational complexity of a set of strings by determining whether it is regular, context-free, decidable, or recognizable.
    \begin{itemize}
         \item {\bf State, prove, and use theorems relating decidability, recognizability, and co-recognizability.}
         \item {\bf Prove that a language is decidable or recognizable by defining and analyzing a Turing machines with appropriate properties.}

   \end{itemize}
   \item  Describe several variants of Turing machines and informally explain why they are equally expressive.
   \begin{itemize}
   \item {\bf Define an enumerator}
   \item {\bf Define nondeterministic Turing machines}
   \item {\bf Use high-level descriptions to define and trace machines (Turing machines and enumerators)}
   \item {\bf Apply dovetailing in high-level definitions of machines}
   \end{itemize}
\end{itemize}
\end{itemize}

\vspace{-20pt}

\subsubsection*{TODO:}
\begin{list}{\itemsep-10pt}
   \item Review Quiz 8 on PrairieLearn (http://us.prairielearn.com), due 3/5/2025
   \item Review Quiz 9 on PrairieLearn (http://us.prairielearn.com), due 3/12/2025
   \item Homework 6 submitted via Gradescope (https://www.gradescope.com/), due 3/13/2025
   \item Project submitted via Gradescope (https://www.gradescope.com/), due 3/19/2025
\end{list}

\newpage

\section*{Monday: Mapping reductions and recognizability}

\input{../activity-snippets/day23.tex}
    
\newpage
\subsection*{Wednesday: More mapping reductions}

\input{../activity-snippets/day24.tex}


\vfill
\subsection*{Friday: Other models of computation}
\input{../activity-snippets/day25.tex}

\newpage
\end{document}