\input{../../resources/lesson-head.tex}

\subsection*{Week 4 at a glance}

\subsubsection*{Textbook reading: Section 1.4, 2.2, 2.1.}

\vspace{-15pt}

Before Monday, read Introduction to Section 1.4 (page 77) which introduces nonregularity.

Before Wednesday, read Definition 2.13 (page 111-112) introducing Pushdown Automata.

Before Friday, read Example 2.18 (page 114).

For Week 5 Monday: read Introduction to Section 2.1 (pages 101-102).

\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}

\begin{itemize}
    \item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
    \begin{itemize}
       \item Give examples of sets that are regular (and prove that they are).
       \begin{itemize}
          \item {\bf State the definition of the class of regular languages}
          \item {\bf Explain the limits of the class of regular languages}
          \item {\bf Identify some regular sets and some nonregular sets}
       \end{itemize}
       \item Use precise notation to formally define the state diagram of a PDA
       \item Use clear English to describe computations of PDA informally.
       \begin{itemize}
           \item {\bf Define push-down automata informally and formally}
           \item {\bf State the formal definition of a PDA}
           \item {\bf Trace the computation(s) of a PDA on a given string using its state diagram}
           \item {\bf Determine if a given string is in the language recognized by a PDA}
           \item {\bf Translate between a state diagram and a formal deﬁnition of a PDA}
           \item {\bf Determine the language recognized by a given PDA}
        \end{itemize}

    \end{itemize}

    \item Know, select and apply appropriate computing knowledge and problem-solving techniques. 
    \begin{itemize}
    \item Apply classical techniques including pumping lemma, determinization, diagonalization, and reduction to analyze the complexity of languages and problems.
    \begin{itemize}
        \item {\bf Justify why the Pumping Lemma is true.}
        \item {\bf Use the pumping lemma to prove that a given language is not regular.}
    \end{itemize}
    \end{itemize}
\end{itemize}

\vspace{-20pt}

\subsubsection*{TODO:}
\begin{list}{\itemsep-10pt}
   \item Schedule your Test 1 Attempt 1, Test 2 Attempt 1, Test 1 Attempt 2, and Test 2 Attempt 2 times 
   at PrairieTest (http://us.prairietest.com)
   \item Review Quiz 3 on PrairieLearn (http://us.prairielearn.com), due 1/29/2025
   \item Homework 2 submitted via Gradescope (https://www.gradescope.com/), due 1/30/2025
   \item Review Quiz 4 on PrairieLearn (http://us.prairielearn.com), due 2/5/2025
\end{list}

\newpage
\subsection*{Monday: Pumping Lemma}

\input{../activity-snippets/day9.tex}

\newpage
\subsection*{Wednesday: Proving nonregularity, and beyond}

\input{../activity-snippets/day10.tex}
    
\newpage
\subsection*{Friday: Pushdown Automata}

\input{../activity-snippets/day11.tex}


\end{document}