
\input{../../resources/lesson-head.tex}

\section*{Monday: Mapping reductions and recognizability}

\input{../activity-snippets/week9-monday.tex}
    
\newpage
\subsection*{Wednesday: More mapping reductions}

\input{../activity-snippets/week9-wednesday.tex}


\newpage
\subsection*{Friday: Other models of computation}

\input{../activity-snippets/week9-friday.tex}

\newpage

\subsection*{Week 9 at a glance}

\subsubsection*{Textbook reading: Section 5.3, 5.1, 3.2}

{\it For Monday}: Example 5.26 (page 237)

{\it For Wednesday}: Theorem 5.30 (page 238)

{\it For Friday}: Skim Section 3.2

{\it For Monday of Week 10}: Definition 7.1 (page 276)

\subsubsection*{Make sure you can:}
\begin{itemize}
\item Classify the computational complexity of a set of strings by determining whether it is decidable or undecidable and recognizable or unrecognizable.
\begin{itemize}
   \item State, prove, and use theorems relating decidability, recognizability, and co-recognizability.
   \item Prove that a language is decidable or recognizable by defining and analyzing a Turing machines with appropriate properties.
   \item Define and explain core examples of computational problems, including $A$**, $E$**, $EQ$**, (for ** either DFA or TM) and $HALT_{TM}$
\end{itemize}
\item Use diagonalization to prove that there are 'hard' languages relative to certain models of computation.
\item Use mapping reduction to deduce the complexity of a language by comparing to the complexity of another.
   \begin{itemize}
      \item Explain what it means for one problem to reduce to another
      \item Define computable functions, and use them to give mapping reductions between computational problems
      \item Define and explain $A_{TM}$ and $HALT_{TM}$
      \item Build and analyze mapping reductions between computational problems
      \item Distinguish between computability and complexity
      \item Articulate motivating questions of complexity
      \item Use appropriate reduction (e.g. mapping, Turing, polynomial-time) to deduce the complexity of a language by comparing to the complexity of another.
   \end{itemize}
\item  Describe several variants of Turing machines and informally explain why they are equally expressive.
   \begin{itemize}
   \item Define an enumerator
   \item Describe the language enumerated by an enumerator
   \item Use high-level descriptions to define and trace machines (Turing machines and enumerators)
   \item Apply dovetailing in high-level definitions of machines
   \item Define nondeterministic Turing machines
   \item State and use the Church-Turing thesis
   \end{itemize}
\end{itemize}


\subsubsection*{TODO:}
\begin{list}
   {\itemsep2pt}
   \item Student Evaluations of Teaching forms: Evaluations are open for completion anytime BEFORE 8AM on Saturday, March 16.
   Access your SETs from the Evaluations site
   \begin{quote}
        \url{https://academicaffairs.ucsd.edu/Modules/Evals}
   \end{quote}
   You will separately evaluate each of your listed instructors for each enrolled course. 

   **NEW** WINTER 2024 SET INCENTIVE LOTTERY: In Winter 2024, students who complete all of their student 
   evaluation forms for their undergraduate course will be entered into a lottery to win one of 
   5 \$100 Visa gift cards! To be entered into the lottery, students must complete at 
   least one instructor evaluation for EACH of their undergraduate courses. 
   They will be automatically entered when they have completed an instructor evaluation for 
   all of their undergraduate courses.

   \item Review quizzes based on class material each day.
   \item Test this Friday in Discussion section.
   \item Homework assignment 5 due next Thursday.
\end{list}

\end{document}
