\input{../../resources/lesson-head.tex}

\subsection*{Week 6 at a glance}

\vspace{-10pt}

\subsubsection*{Textbook reading: Chapter 3}

\vspace{-10pt}

Before Monday, Page 165-166 Introduction to Section 3.1.

\vspace{-10pt}

Before Wednesday, Example 3.9 on page 173.

\vspace{-10pt}

Before Friday, Page 184-185 Terminology for describing Turing machines.

\vspace{-10pt}

Week 7 Monday: No class, in observance of Veterans Day. For Week 7 Wednesday: Introduction to Chapter 4.

\vspace{-20pt}

\subsubsection*{We will be learning and practicing to:}
\vspace{-20pt}

\begin{itemize}
    \item Clearly and unambiguously communicate computational ideas using appropriate formalism. Translate across levels of abstraction.
    \begin{itemize}
        \item Use precise notation to formally define the state diagram of a Turing machine
        \item Use clear English to describe computations of Turing machines informally.
        \begin{itemize}
                \item {\bf Motivate the definition of a Turing machine}
                \item {\bf Trace the computation of a Turing machine on given input}
                \item {\bf Describe the language recognized by a Turing machine}
                \item {\bf Determine if a Turing machine is a decider}
                \item {\bf Given an implementation-level description of a Turing machine}
                \item {\bf Use high-level descriptions to define and trace Turing machines}
                \item {\bf Apply dovetailing in high-level definitions of machines}
         \end{itemize}
       \item Give examples of sets that are recognizable and decidable (and prove that they are).
       \begin{itemize}
          \item {\bf State the definition of the class of recognizable languages}
          \item {\bf State the definition of the class of decidable languages}
          \item {\bf State the definition of the class of co-recognizable languages}
       \end{itemize}
    \end{itemize}
    \item Know, select and apply appropriate computing knowledge and problem-solving techniques. 
    Reason about computation and systems.
    \item Describe and prove closure properties of classes of languages under certain operations.
    \begin{itemize}
        \item {\bf Apply a general construction to create a new Turing machine from an example one.}
        \item {\bf Formalize a general construction from an informal description of it.}
        \item {\bf Use general constructions to prove closure properties of the class of decidable langugages.}
        \item {\bf Use general constructions to prove closure properties of the class of recognizable langugages.}
    \end{itemize}
\end{itemize}

\vspace{-20pt}

\subsubsection*{TODO:}
\begin{list}{\itemsep -20pt}
   \item Review Quiz 6 on PrairieLearn (http://us.prairielearn.com), complete by Sunday 11/11/2024
   \item Homework 4 submitted via Gradescope (https://www.gradescope.com/), due Tuesday 10/12/2024
\end{list}

\newpage

\section*{Monday: Descriptions of Turing machines}

\input{../activity-snippets/week6-monday.tex}
    
\newpage
\subsection*{Wednesday: Recognizable and decidable languages}

\input{../activity-snippets/week6-wednesday.tex}


\newpage
\subsection*{Friday: Closure for the classes of recognizable and decidable languages}

\input{../activity-snippets/week6-friday.tex}


\end{document}
