\input{../../resources/lesson-head.tex}

\section*{Before we start}

We want you to be successful. 

We will work together to build an 
environment in CSE 105 that supports your learning
in a way that respects your
perspectives, experiences, and identities (including race, ethnicity, heritage, gender, sex, 
class, sexuality, religion, ability, age, educational background, etc.).  
Our goal is for you to  engage
with interesting and challenging concepts and 
feel comfortable exploring, asking questions, and thriving.

If you or someone you know is suffering from food and/or housing insecurities 
there are UCSD resources here to help:

Basic Needs Office: \href{https://basicneeds.ucsd.edu/}{https://basicneeds.ucsd.edu/}

Triton Food Pantry (in the old Student Center)
is free and anonymous, and includes produce: 

\href{https://www.facebook.com/tritonfoodpantry/}{https://www.facebook.com/tritonfoodpantry/}

Mutual Aid UCSD: \href{https://mutualaiducsd.wordpress.com/}{https://mutualaiducsd.wordpress.com/}

Financial aid resources, the possibility of emergency grant funding, and off-campus housing referral 
resources are available: see your College Dean of Student Affairs.

If you find yourself in an uncomfortable situation, ask for help. 
We are committed to upholding University policies regarding nondiscrimination, sexual violence and sexual harassment:
Counseling and Psychological Services (CAPS) at 858 5343755 or \href{http://caps.ucsd.edu}{http://caps.ucsd.edu}; 
OPHD at (858) 534-8298, ophd@ucsd.edu , \href{http://ophd.ucsd.edu}{http://ophd.ucsd.edu};
CARE at Sexual Assault Resource Center at 858 5345793 sarc@ucsd.edu \href{http://care.ucsd.edu}{http://care.ucsd.edu}.


Please reach out (minnes@ucsd.edu) if you need support with extenuating circumstances.

\vfill

\section*{Introductions}
Class website: \href{https://canvas.ucsd.edu/courses/51649/}{https://canvas.ucsd.edu/courses/51649/}


Instructor: Prof. Mia Minnes {\tiny{"Minnes" rhymes with Guinness}}, minnes@ucsd.edu, 
\href{http://cseweb.ucsd.edu/~minnes}{http://cseweb.ucsd.edu/~minnes}


Our team: One instructor + two TAs and eleven tutors + all of you

Fill in contact info for students around you, if you'd like:

\vfill

\begin{comment}
On a typical week: {\bf MWF} Lectures + review quizzes, {\bf Tu} Homework due, {\bf WF} Discussion.
Office hours (hosted by instructors and TAs and tutors, where you can come to talk 
about course concepts and ask for help as you work through sample problems) and Q+A on Piazza available throughout the week.
All dates are on \href{https://canvas.ucsd.edu/}{Canvas (click for link)} and details are on
 \href{https://canvas.ucsd.edu/courses/51649/}{the Syllabus page}.


There are lots of great reasons to have a laptop, tablet, or phone open during class. You might be taking notes, 
getting a photo of an important moment on the board, trying out a construction that we're developing together, working 
on the review quiz, and so on. 
The main issue with screens and technology in the classroom isn't your own distraction (which is your responsibility to manage), 
it's the distraction of other students. Anyone sitting behind you cannot help but have your screen in their field of view. 
Having distracting content on your screen can really harm their learning experience.

With this in mind, the device policy for the course is that if you have a screen open, you either:

Have only content onscreen that's directly related to the current lecture.
Have unrelated content open and sit in one of the back two rows of the room to mitigate the effects on other students. 
\end{comment}

\newpage Welcome to CSE 105: Introduction to Theory of Computation in Winter 2024!

\section*{CSE 105's Big Questions}
\begin{itemize}
   \item What problems are computers capable of solving?
   \item What resources are needed to solve a problem?
   \item Are some problems harder than others?
\end{itemize}

In this context, a {\bf problem} is defined as: ``Making a decision or computing a value based on some input"

Consider the following problems: 
\begin{itemize}
   \item Find a file on your computer
   \item Determine if your code will compile
   \item Find a run-time error in your code
   \item Certify that your system is un-hackable
\end{itemize}

Which of these is hardest?

\vfill

In Computer Science, we operationalize ``hardest'' as ``requires most resources'', where
resources might be memory, time, parallelism, randomness, power, etc.

To be able to compare ``hardness'' of problems, we use a consistent description of problems

{\bf Input}: String

{\bf Output}: Yes/ No, where Yes means that the input string matches the pattern or property described by the problem.


\newpage
\section*{Monday: Terminology and Notation}

\input{../activity-snippets/week1-monday.tex}
    
\newpage
\subsection*{Review: Week 1 Monday}
\begin{enumerate}
\item Please complete the beginning of the quarter survey linked from our \#FinAid
Assignment on Canvas https://canvas.ucsd.edu/courses/51649/assignments/734784?wrap=1
\item We want you to be familiar with class policies and procedures so you are ready to have a successful quarter. 
Please take a look at the syllabus on our class website https://canvas.ucsd.edu/courses/45073
and answer the questions about it on Gradescope.
\end{enumerate}

A summary of the terminology we will use is on page 16 in the textbook.

{\bf Pre class reading for next time}: Example 1.51, Definition 1.52

{\it Notice}: we are jumping to Section 1.3 Regular Expressions and then will come back 
to Section 1.1 Finite Automata on Friday.


\newpage
\subsection*{Week 1 Wednesday}

\input{../activity-snippets/week1-wednesday.tex}

\newpage
\subsection*{Review: Week 1 Wednesday}

Please complete the review quiz questions on \href{http://gradescope.com}{Gradescope} about 
strings, languages, and regular expressions.


Recall: Review quizzes based on class material are assigned each day. 
These quizzes will help you track and confirm your understanding of the concepts and examples 
we work in class. Quizzes can be submitted on Gradescope as many times (with no penalty) as 
you like until the quiz deadline: the three quizzes each week are all due on Friday (with no penalty 
late submission open until Sunday).

The definition of the union, concatenation, and star operations for languages is given 
as Definition 1.23 on page 44 and a useful example is Example 1.24.

{\bf Pre class reading for next time}: Figure 1.4, Definition 1.5

{\it Notice}: On Friday we are going back to the start of Chapter 1 and will be discussing finite 
automata and their computations.

\newpage
\subsection*{Week 1 Friday}

\input{../activity-snippets/week1-friday.tex}


\newpage

\subsection*{Review: Week 1 Friday}

Please complete the review quiz questions on \href{http://gradescope.com}{Gradescope} about 
DFA and the strings they accept and reject.

The definition of the union, concatenation, and star operations for languages is given 
as Definition 1.23 on page 44 and a useful example is Example 1.24.


The first homework assignment for CSE 105 this quarter is due next week. 
Find it on the class website: in the Calendar section of the website, click on the link under "Due Dates" for Week 2, or find
the detailed assignment in the Assignments section of the website.
Due: 4/11/23 at 5pm (no penalty late submission until 8am next morning)

We encourage you to work on homework in groups of up to three CSE 105 classmates. 
To find group members: reach out to people sitting around you in class, in discussion section, 
or during office hours. The pre-survey also asks if you want help finding group members: 
the CSE 105 instructional team can help you connect with other students. Working within the 
campus safety guidelines, you may choose to meet with your group mates in person or remotely. 
We highly recommend meeting synchronously so that you can work through the homework problems *together*. 

{\bf Pre class reading for next time}: pages 41-43 (figures 1.18, 1.19, 1.20)

\end{document}