\input{../../resources/assignment-head.tex}

\title{HW4CSE105W24: Homework assignment 4}
\date{Due: February 29th at 5pm (no penalty late submission until 8am next morning), via Gradescope}

\begin{document}
\maketitle
\thispagestyle{fancy}

{\bf In this assignment,}
You will  practice analyzing, designing, and working with Turing machines.
You will use general constructions and specific machines to explore the classes 
of recognizable and decidable languages. 
You will explore various ways to encode machines as strings so that 
computational problems can be recognized.

{\bf Resources}: To review the topics 
for this assignment, see the class material from Weeks 5-7.
We will post frequently asked questions and our answers to them in a 
pinned Piazza post.

{\bf Reading and extra practice problems}:  
Sipser Sections 3.1, 3.3, 4.1
Chapter 3 exercises 3.1, 3.2, 3.5, 3.8.
Chapter 4 exercises 4.1, 4.2, 4.3, 4.4, 4.5.

\instructions

You will submit this assignment via Gradescope
(\href{https://www.gradescope.com}{https://www.gradescope.com}) 
in the assignment called ``hw4CSE105W24''.

{\bf Assigned questions}
\begin{enumerate}[wide, labelwidth=!, labelindent=0pt]


%%%%%%%%%%% PROBLEM 1 %%%%%%%%%%%
\item\textbf{Classifying languages} (10 points):
Our first example of a more complicated Turing machine was of a Turing machine 
that recognized the language $\{w \# w \mid w \in\{0,1\}^*\}$, which we 
know is not context-free. The language
\[
    \{0^n 1^n 2^n \mid n \geq 0\}
\]
is also not context-free. 

\begin{enumerate}
    \item\gradeCorrectFirst Give an implementation-level description of a Turing machine 
    that recognizes this language.
    \item\gradeCompleteFirst Draw a state diagram of the Turing machine you gave in part (a) and trace
    the computation of this Turing machine on the input $012$. You may use all 
    our usual conventions for state diagrams of Turing machines 
    (we do not include the node for the reject state $qrej$ and any missing transitions 
    in the state diagram have value $(qrej,\square,R)$; 
    $b \to R$ label means $b \to b, R$ ).
\end{enumerate}


%%%%%%%%%%% PROBLEM 2 %%%%%%%%%%%
\item \textbf{Deciders, Recognizers, Decidability, and Recognizability} (15 points):
For this question, consider the alphabet $\Sigma = \{0,1\}$.
\begin{enumerate}
\item[(a)]\gradeCorrect Give an example of a finite, nonempty language over $\Sigma$ and 
two different Turing machines that recognize it: one that is a decider and one that is not.
A complete solution will include a precise definition for your example language, 
along with {\bf both} a state diagram and an implementation-level description 
of each Turing machines, along with a brief explanation of why each of them recognizes
the language and why one is a decider and there other is not.

\item[(b)]\gradeCorrect True or false: There is a Turing machine that is not a decider that recognizes 
the empty set. A complete solution will include a witness Turing machine (given by 
state diagram or implementation-level description or high-level description) and a justification 
for why it's not a decider and why it does not accept any strings, or a complete and correct
justification for why there is no such Turing machine.

\item[(c)]\gradeCorrect True or false: There is a Turing machine that is not a decider that recognizes 
the set of all string $\Sigma^*$.  A complete solution will include a witness Turing machine 
(given by 
state diagram or implementation-level description or high-level description) and a justification 
for why it's not a decider and why it accept each string over $\{0,1\}$, or a complete and correct
justification for why there is no such Turing machine.
\end{enumerate}

%%%%%%%%%%% PROBLEM 3 %%%%%%%%%%%
\item\textbf{Closure} (15 points):
Suppose $M$ is a Turing machine over the alphabet $\{0,1\}$. 
Let $s_1, s_2, \ldots$ be a list of all strings in 
$\{0,1\}^*$ in string (shortlex) order.
We define a new Turing machine 
by giving its high-level description as follows: 
\begin{align*}
   M_{new} &= ``\text{On input }w:\\
    &\text{1. For $n = 1, 2, \ldots$}\\
    &\text{2.~~~For $j = 1, 2, \ldots n$} \\
    &\text{3.~~~For $k = 1, 2, \ldots, n$} \\
    &\text{4.~~~~~~Run the computation of $M$ on $s_jws_k$}\\
    &\text{5.~~~~~~If it accepts, accept.}\\
    &\text{6.~~~~~~If it rejects, go to the next iteration of the loop"}\\
\end{align*}

Recall the definitions we have: 
For languages $L_1, L_2$ over the alphabet $\Sigma = \{0,1\}$, we have the 
associated sets of strings
\[
   SUBSTRING(L_1) = \{ w \in \Sigma^* ~|~ \text{there exist } a,b \in \Sigma^* \text{ such that } awb \in L_1\}
\]
and 
\[
   L_1 \circ L_2 = \{ w \in \Sigma^* ~|~ w = uv \text{ for some strings } u \in L_1 \text{ and } v \in L_2 \}
\]
We say that self-set-wise concatenation of the set $L_1$ is $L_1 \circ L_1$.


{\it Note: there was a bug in the version of this assignment 
that was first released.}

\begin{enumerate}
\item[(a)]\gradeComplete Prove that this Turing machine construction 
{\bf cannot} be used to prove that the
class of decidable languages over $\{0,1\}$ is closed under {\bf either} of the above operations
($SUBSTRING$ or self-set-wise concatenation).
A complete answer will give a counterexample or general description why the construction doesn't work for 
both operations.
\item[(b)] \gradeCorrect Prove that this Turing machine construction cannot be used to prove that the
class of recognizable languages over $\{0,1\}$ is closed under the $SUBSTRING$ set operation. In particular, 
give a counterexample of a specific language $L_1$ and Turing machine $M_1$ recognizing it where $M_{new}$ does not
recognize $SUBSTRING(L_1)$.
\item[(c)] \gradeComplete Define a new construction by slightly modifiying this one that can be used to prove  that the
class of recognizable languages over $\{0,1\}$ is closed under $SUBSTRING$. Justify that 
your construction works. The proof of correctness for the closure claim can be structured like: 
``Let $L_1 $ be a recognizable language over $\{0,1\}$ and 
assume we are given a Turing machine $M_1$ so that $L(M_1) = L_1$. Consider the new Turing machine 
$M_{new}$ defined above. We will show that $L(M_{new}) = SUBSTRING(L_1) $... {\it complete the proof
by proving subset inclusion in two directions, by tracing the relevant Turing machine computations}''
\end{enumerate}

%%%%%%%%%%% PROBLEM 4 %%%%%%%%%%%
\item \textbf{Computational problems} (10 points):
Recall the definitions of some example computational problems from class

\hspace{-30pt}
    \begin{tabular}{|lcl|}
    \hline
    \multicolumn{3}{|l|}{{\bf  Acceptance problem} } \\
    & & \\
    \ldots for DFA & $A_{DFA}$ & $\{ \langle B,w \rangle \mid  \text{$B$ is a  DFA that accepts input 
    string $w$}\}$ \\
    \ldots for NFA & $A_{NFA}$ & $\{ \langle B,w \rangle \mid  \text{$B$ is a  NFA that accepts input 
    string $w$}\}$ \\
    \ldots for regular expressions & $A_{REX}$ & $\{ \langle R,w \rangle \mid  \text{$R$ is a  regular
    expression that generates input string $w$}\}$ \\
    \ldots for CFG & $A_{CFG}$ & $\{ \langle G,w \rangle \mid  \text{$G$ is a context-free grammar 
    that generates input string $w$}\}$ \\
    \ldots for PDA & $A_{PDA}$ & $\{ \langle B,w \rangle \mid  \text{$B$ is a PDA that accepts input string $w$}\}$ \\
    & & \\
    \hline
    \multicolumn{3}{|l|}{{\bf Language emptiness  testing} } \\
    & & \\
    \ldots for DFA & $E_{DFA}$ & $\{ \langle A \rangle \mid  \text{$A$ is a  DFA and  $L(A) = \emptyset$\}}$ \\
    \ldots for NFA & $E_{NFA}$ & $\{ \langle A\rangle \mid  \text{$A$ is a NFA and  $L(A) = \emptyset$\}}$ \\
    \ldots for regular expressions & $E_{REX}$ & $\{ \langle R \rangle \mid  \text{$R$ is a  regular
    expression and  $L(R) = \emptyset$\}}$ \\
    \ldots for CFG & $E_{CFG}$ & $\{ \langle G \rangle \mid  \text{$G$ is a context-free grammar 
    and  $L(G) = \emptyset$\}}$ \\
    \ldots for PDA & $E_{PDA}$ & $\{ \langle A \rangle \mid  \text{$A$ is a PDA and  $L(A) = \emptyset$\}}$ \\
    & & \\
    \hline
    \multicolumn{3}{|l|}{{\bf Language equality testing} } \\
    & & \\
    \ldots for DFA & $EQ_{DFA}$ & $\{ \langle A, B \rangle \mid  \text{$A$ and $B$ are DFAs and  $L(A) =L(B)$\}}$\\
    \ldots for NFA & $EQ_{NFA}$ & $\{ \langle A, B \rangle \mid  \text{$A$ and $B$ are NFAs and  $L(A) =L(B)$\}}$\\
    \ldots for regular expressions & $EQ_{REX}$ & $\{ \langle R, R' \rangle \mid  \text{$R$ and $R'$ are regular
    expressions and  $L(R) =L(R')$\}}$\\
    \ldots for CFG & $EQ_{CFG}$ & $\{ \langle G, G' \rangle \mid  \text{$G$ and $G'$ are CFGs and  $L(G) =L(G')$\}}$ \\
    \ldots for PDA & $EQ_{PDA}$ & $\{ \langle A, B \rangle \mid  \text{$A$ and $B$ are PDAs and  $L(A) =L(B)$\}}$ \\
    \hline
    \end{tabular}

\begin{enumerate}
    \item[(a)] \gradeComplete Pick five of the computational problems above and give 
    examples (preferably different from the ones we talked about in class) of strings that are
    in each of the corresponding languages. Remember to use the 
    notation $\langle \cdots \rangle$ to denote the string encoding of relevant objects.
    {\it Extension, not for credit:} Explain why it's hard to write a specific string of 
    $0$s and $1$s and make a claim about membership in one of these sets.
    \item[(b)] \gradeComplete Computational problems can also be defined about Turing machines.
    Consider the two high-level descriptions of Turing machines below.
    Reverse-engineer them to define the computational problem that is being
    recognized, where $L(M_{DFA})$ is the language corresponding to this computational
    problem about DFA and $L(M_{TM})$ is the language corresponding to this computational
    problem about Turing machines. {\it Hint}: the computational problem is not acceptance,
    language emptiness, or language equality (but is related to one of them).

    Let $s_1, s_2, \ldots$ be a list of all strings in 
    $\{0,1\}^*$ in string (shortlex) order. Consider the following Turing machines
    \begin{align*}
        M_{DFA} &= ``\text{On input $\langle D \rangle$ where $D$ is a DFA}:\\
         &\text{1. for $i=1, 2, 3, \ldots$} \\
         &\text{2.~~~ Run $D$ on $s_i$} \\
         &\text{3.~~~~If it accepts, accept.}\\
         &\text{4.~~~~If it rejects, go to the next iteration of the loop"}\\
     \end{align*}
     and
     \begin{align*}
        M_{TM} &= ``\text{On input $\langle T \rangle$ where $T$ is a Turing machine}:\\
         &\text{1. for $i=1, 2, 3, \ldots$} \\
         &\text{2.~~~ Run $T$ for $i$ steps on each input $s_1, s_2, \ldots, s_i$ in turn} \\
         &\text{3.~~~~If $T$ has accepted any of these, accept.}\\
         &\text{4.~~~~Otherwise, go to the next iteration of the loop"}\\
     \end{align*}
\end{enumerate}


\end{enumerate}
\end{document}