\input{../../resources/assignment-head.tex}

\title{HW5CSE105F24: Homework assignment 5}
\date{Due: November 19, 2024 at 5pm, via Gradescope}


\begin{document}
\maketitle
\thispagestyle{fancy}

{\bf In this assignment,}

You will  practice analyzing, designing, and working with Turing machines.
You will use general constructions and specific machines to explore the classes 
of recognizable and decidable languages. 
You will explore various ways to encode machines as strings so that 
computational problems can be recognized.

{\bf Resources}: To review the topics 
for this assignment, see the class material from Weeks 6 and 7.
We will post frequently asked questions and our answers to them in a 
pinned Piazza post. 

{\bf Reading and extra practice problems}:  
Sipser Sections 3.1, 3.3, 4.1
Chapter 3 exercises 3.1, 3.2, 3.5, 3.8.
Chapter 4 exercises 4.1, 4.2, 4.3, 4.4, 4.5.

\instructions

You will submit this assignment via Gradescope
(\href{https://www.gradescope.com}{https://www.gradescope.com}) 
in the assignment called ``hw5CSE105F24''.

{\bf Assigned questions}
\begin{enumerate}[wide, labelwidth=!, labelindent=0pt]

%%%%%%%%%%% PROBLEM 1 %%%%%%%%%%%
\item\textbf{Classifying languages} (10 points):
Our first example of a more complicated Turing machine was of a Turing machine 
that recognized the language $\{w \# w \mid w \in\{0,1\}^*\}$, which we 
know is not context-free. Let's call that Turing machine $M_0$. The language
\[
    L = \{ww \mid w \in \{0,1\}^*\}
\]
is also not context-free. 

\begin{enumerate}
    \item\gradeCorrectFirst Choose an example string of length $4$ in $L$ that is in {\bf not} in $\{w \# w \mid w \in\{0,1\}^*\}$ and describe the computation of the Turing machine $M_0$ on your example string. 
    Include the contents of the  tape, the state of the machine, and the location of the read/write head at each step in the computation.
    \item\gradeCompleteFirst Explain why the Turing machine from the textbook 
    and class that recognized $\{w \# w \mid w \in\{0,1\}^*\}$ does 
    not recognize $\{ww \mid w \in \{0,1\}^*\}$. Use your example to explain why $M_0$ doesn't recognize $L$.
    \item\gradeComplete Explain how you would change $M_0$ to get a 
    new Turing machine that does recognize $L$. Describe this new Turing machine using both an implementation-level definition and a state diagram of the Turing machine. You may use all 
    our usual conventions for state diagrams of Turing machines 
    (we do not include the node for the reject state $qrej$ and any missing transitions 
    in the state diagram have value $(qrej,\square,R)$; 
    $b \to R$ label means $b \to b, R$ ).
\end{enumerate}

%%%%%%%%%%% PROBLEM 2 %%%%%%%%%%%
\item\textbf{Closure} (18 points):
Suppose $M$ is a Turing machine over the alphabet $\{0,1\}$. 
Let $s_1, s_2, \ldots$ be a list of all strings in 
$\{0,1\}^*$ in string (shortlex) order.
We define a new Turing machine 
by giving its high-level description as follows: 
\begin{align*}
   M_{new} &= ``\text{On input }w:\\
    &\text{1. For $n = 1, 2, \ldots$}\\
    &\text{2.~~~For $j = 1, 2, \ldots n$} \\
    &\text{3.~~~For $k = 1, 2, \ldots, n$} \\
    &\text{4.~~~~~~Run the computation of $M$ on $s_jws_k$}\\
    &\text{5.~~~~~~If it accepts, accept.}\\
    &\text{6.~~~~~~If it rejects, go to the next iteration of the loop"}\\
\end{align*}

Recall the definitions we have:   For each language $L$ over the alphabet $\Sigma_1 = \{0,1\}$, we have the 
associated sets of strings
  $$SUBSTRING(L) = \{ w \in \Sigma_1^* ~|~ \text{there exist } x,y \in \Sigma_1^* \text{ such that } xwy \in L\}$$
and 
  $$EXTEND(L) = \{ w \in \Sigma_1^* ~|~ w = uv \text{ for some strings } u \in L \text{ and } v \in \Sigma_1^* \}$$

\begin{enumerate}
\item[(a)]\gradeCorrect Prove that this Turing machine construction 
{\bf cannot} be used to prove that the
class of decidable languages over $\{0,1\}$ is closed under the 
$EXTEND$ operation.
A complete and correct answer will give a counterexample which is a set $A$ over $\Sigma_1$ that  is decidable, along with a definition of Turing machine $M_A$ 
that decides $A$ 
(with a justification why this Turing machine accepts all 
strings in $A$ and rejects all strings not in $A$), and then either a description of the language of $M_{new}$ that results when setting the Turing machine $M = M_A$ and an explanation why $L(M_{new}) \neq EXTEND(A)$ or 
a description why $M_{new}$ is not a decider and therefore can't witness that $EXTEND(A)$ is decidable.

\item[(b)] \gradeCorrect Prove that this Turing machine construction cannot be used to prove that the
class of recognizable languages over $\{0,1\}$ is closed under the $SUBSTRING$ set operation. A complete and correct answer will 
give a counterexample of a specific language $B$ and Turing machine $M_B$ 
recognizing it
(with a justification why this Turing machine accepts all and only strings in $B$), and then a description of the language of $M_{new}$ that results when 
setting the Turing machine $M = M_B$ and an explanation why $L(M_{new}) \neq SUBSTRING(B)$

\item[(c)] \gradeComplete Define a new construction by slightly modifiying this one that can be used to prove  that the
class of recognizable languages over $\{0,1\}$ is closed under $SUBSTRING$. Justify that 
your construction works. The proof of correctness for the closure claim can be structured like: 
``Let $L_1 $ be a recognizable language over $\{0,1\}$ and 
assume we are given a Turing machine $M_1$ so that $L(M_1) = L_1$. Consider the new Turing machine 
$M_{new}$ defined above. We will show that $L(M_{new}) = SUBSTRING(L_1) $... {\it complete the proof
by proving subset inclusion in two directions, by tracing the relevant Turing machine computations}''

\item[(d)] \gradeComplete Prove that the class of recognizable languages over 
$\{0,1\}$ is closed under $EXTEND$.
\end{enumerate}

%%%%%%%%%%% PROBLEM 3 %%%%%%%%%%%
\item \textbf{Computational problems} (12 points):
Recall the definitions of some example computational problems from class

\hspace{-30pt}
    \begin{tabular}{|lcl|}
    \hline
    \multicolumn{3}{|l|}{{\bf  Acceptance problem} } \\
    & & \\
    \ldots for DFA & $A_{DFA}$ & $\{ \langle B,w \rangle \mid  \text{$B$ is a  DFA that accepts input 
    string $w$}\}$ \\
    \ldots for NFA & $A_{NFA}$ & $\{ \langle B,w \rangle \mid  \text{$B$ is a  NFA that accepts input 
    string $w$}\}$ \\
    \ldots for regular expressions & $A_{REX}$ & $\{ \langle R,w \rangle \mid  \text{$R$ is a  regular
    expression that generates input string $w$}\}$ \\
    \ldots for CFG & $A_{CFG}$ & $\{ \langle G,w \rangle \mid  \text{$G$ is a context-free grammar 
    that generates input string $w$}\}$ \\
    \ldots for PDA & $A_{PDA}$ & $\{ \langle B,w \rangle \mid  \text{$B$ is a PDA that accepts input string $w$}\}$ \\
    & & \\
    \hline
    \multicolumn{3}{|l|}{{\bf Language emptiness  testing} } \\
    & & \\
    \ldots for DFA & $E_{DFA}$ & $\{ \langle A \rangle \mid  \text{$A$ is a  DFA and  $L(A) = \emptyset$\}}$ \\
    \ldots for NFA & $E_{NFA}$ & $\{ \langle A\rangle \mid  \text{$A$ is a NFA and  $L(A) = \emptyset$\}}$ \\
    \ldots for regular expressions & $E_{REX}$ & $\{ \langle R \rangle \mid  \text{$R$ is a  regular
    expression and  $L(R) = \emptyset$\}}$ \\
    \ldots for CFG & $E_{CFG}$ & $\{ \langle G \rangle \mid  \text{$G$ is a context-free grammar 
    and  $L(G) = \emptyset$\}}$ \\
    \ldots for PDA & $E_{PDA}$ & $\{ \langle A \rangle \mid  \text{$A$ is a PDA and  $L(A) = \emptyset$\}}$ \\
    & & \\
    \hline
    \multicolumn{3}{|l|}{{\bf Language equality testing} } \\
    & & \\
    \ldots for DFA & $EQ_{DFA}$ & $\{ \langle A, B \rangle \mid  \text{$A$ and $B$ are DFAs and  $L(A) =L(B)$\}}$\\
    \ldots for NFA & $EQ_{NFA}$ & $\{ \langle A, B \rangle \mid  \text{$A$ and $B$ are NFAs and  $L(A) =L(B)$\}}$\\
    \ldots for regular expressions & $EQ_{REX}$ & $\{ \langle R, R' \rangle \mid  \text{$R$ and $R'$ are regular
    expressions and  $L(R) =L(R')$\}}$\\
    \ldots for CFG & $EQ_{CFG}$ & $\{ \langle G, G' \rangle \mid  \text{$G$ and $G'$ are CFGs and  $L(G) =L(G')$\}}$ \\
    \ldots for PDA & $EQ_{PDA}$ & $\{ \langle A, B \rangle \mid  \text{$A$ and $B$ are PDAs and  $L(A) =L(B)$\}}$ \\
    \hline
    \end{tabular}

\begin{enumerate}
    \item[(a)] \gradeComplete Pick five of the computational problems above and give 
    examples (preferably different from the ones we talked about in class) of strings that are
    in each of the corresponding languages. Remember to use the 
    notation $\langle \cdots \rangle$ to denote the string encoding of relevant objects.
    {\it Extension, not for credit:} Explain why it's hard to write a specific string of 
    $0$s and $1$s and make a claim about membership in one of these sets.
    \item[(b)] \gradeComplete Computational problems can also be defined about Turing machines.
    Consider the two high-level descriptions of Turing machines below.
    Reverse-engineer them to define the computational problem that is being
    recognized, where $L(M_{DFA})$ is the language corresponding to this computational
    problem about DFA and $L(M_{TM})$ is the language corresponding to this computational
    problem about Turing machines. {\it Hint}: the computational problem is not acceptance,
    language emptiness, or language equality (but is related to one of them).

    Let $s_1, s_2, \ldots$ be a list of all strings in 
    $\{0,1\}^*$ in string (shortlex) order. Consider the following Turing machines
    \begin{align*}
        M_{DFA} &= ``\text{On input $\langle D \rangle$ where $D$ is a DFA}:\\
         &\text{1. for $i=1, 2, 3, \ldots$} \\
         &\text{2.~~~ Run $D$ on $s_i$} \\
         &\text{3.~~~~If it accepts, accept.}\\
         &\text{4.~~~~If it rejects, go to the next iteration of the loop"}\\
     \end{align*}
     and
     \begin{align*}
        M_{TM} &= ``\text{On input $\langle T \rangle$ where $T$ is a Turing machine}:\\
         &\text{1. for $i=1, 2, 3, \ldots$} \\
         &\text{2.~~~ Run $T$ for $i$ steps on each input $s_1, s_2, \ldots, s_i$ in turn} \\
         &\text{3.~~~~If $T$ has accepted any of these, accept.}\\
         &\text{4.~~~~Otherwise, go to the next iteration of the loop"}\\
     \end{align*}
\end{enumerate}


%%%%%%%%%%% PROBLEM 4 %%%%%%%%%%%
\item\textbf{Computational problems} (10 points):
For each of the following statements, determine if it is true or false. 
Clearly label your choice by starting your solution with True or False and then provide a 
 justification for your answer.

\begin{enumerate}
    \item\gradeCorrect Prove that the language $$\{\langle D \rangle \mid D \text{ is an NFA over $\{0,1\}$ and } L(D) = L(0^*\cup 1^*) \}$$
    is decidable.
    \item\gradeCorrect Prove that the language $$\{\langle R_1, R_2 \rangle \mid R_1, R_2 \text{ are regular expressions over $\{0,1\}$ and } L(R_1) \subseteq L(R_2) \}$$ is decidable.
\end{enumerate}

\end{enumerate}
\end{document}