\input{../../resources/assignment-head.tex}

\title{HW6CSE105F24: Homework assignment 6}
\date{Due: December 3, 2024 at 5pm, via Gradescope}


\begin{document}
\maketitle
\thispagestyle{fancy}

{\bf In this assignment,}

You will  practice analyzing, designing, and working with Turing machines.
You will use general constructions and specific machines to explore the classes 
of recognizable and decidable languages. 
You will explore various ways to encode machines as strings so that 
computational problems can be recognized.

{\bf Resources}: To review the topics 
for this assignment, see the class material from Weeks 8 and 9.
We will post frequently asked questions and our answers to them in a 
pinned Piazza post. 

{\bf Reading and extra practice problems}:  
Sipser Sections 4.2, 5.3, 5.1.
Chapter 4 exercises 4.9, 4.12.
Chapter 5 exercises 5.4, 5.5, 5.6, 5.7. 
Chapter 5 problems 5.22, 5.23, 5.24, 5.28

\instructions

You will submit this assignment via Gradescope
(\href{https://www.gradescope.com}{https://www.gradescope.com}) 
in the assignment called ``hw6CSE105F24''.

{\bf Assigned questions}
\begin{enumerate}[wide, labelwidth=!, labelindent=0pt]

%%%%%%%%%%% PROBLEM 1 %%%%%%%%%%%
\item \textbf{What's wrong with these reductions? (if anything)} (15 points):
Suppose your friends are practicing
coming up with mapping reductions $A \leq_m B$ and their witnessing
functions $f: \Sigma^* \to \Sigma^*$. For each of the following 
attempts, determine if it  has error(s) or is correct.
Do so by labelling each attempt with all and 
only the labels below that apply, and justifying
this labelling.
\begin{itemize}
\item \textit{Error Type 1:} The given function 
can't witness the claimed mapping reduction because there
exists an $x \in A$ such that $f(x) \not\in B$.
\item \textit{Error Type 2:} The given function 
can't witness the claimed mapping reduction because there 
exists an $x \not\in A$ such that $f(x) \in B$.
\item \textit{Error Type 3:} The given function 
can't witness the claimed mapping reduction because the specified
function is not computable.
\item \textit{Correct:} The 
claimed mapping reduction is true and 
is witnessed by the given function.
\end{itemize}

Clearly present your answer by
providing a brief (3-4 sentences or so) justification for 
whether {\bf each} of these labels applies to each example.


\begin{enumerate}
\item\gradeCompleteFirst $A_{\mathrm{TM}} \le_m HALT_{\mathrm{TM}}$ and 
\[
f(x) = \begin{cases}
 \scalebox{.7}{$\langle$ \hspace{-.5cm} \raisebox{-.4cm}{
\begin{tikzpicture}[->,>=stealth',shorten >=1pt, auto, node distance=2cm, semithick]
  \tikzstyle{every state}=[text=black, fill=none]
  \node[initial,state,accepting] (q0)                    {$q_{\mathrm{acc}}$};
 ;
\end{tikzpicture}}
$, \varepsilon \rangle$}  
& \text{if } x = \langle M, w \rangle \text{ for a Turing machine $M$ and string $w$}\\
& \qquad \qquad \text{ and } w \in L(M) \\

\scalebox{.7}{$\langle$ \hspace{-.5cm} \raisebox{-.4cm}{
\begin{tikzpicture}[->,>=stealth',shorten >=1pt, auto, node distance=2cm, semithick]
  \tikzstyle{every state}=[text=black, fill=none]
  \node[initial,state] (q0)                    {$q_0$};
  \node[state,accepting] (qacc) [right of = q0, xshift = 20]{$q_{acc}$};
  \path (q0) edge  [loop above] node {$0, 1, \blank \to R$} (q0)
 ;
\end{tikzpicture}}
$\rangle$} 
& \text{otherwise}
\end{cases} 
\]


\item\gradeComplete $A_{\mathrm{TM}} \le_m EQ_{\mathrm{TM}} $ with 
\[
f(x) = \begin{cases}
 \scalebox{.7}{$\langle$ \hspace{-.5cm} \raisebox{-.4cm}{
\begin{tikzpicture}[->,>=stealth',shorten >=1pt, auto, node distance=2cm, semithick]
  \tikzstyle{every state}=[text=black, fill=none]
  \node[initial,state,accepting] (q0)                    {$q_{\mathrm{acc}}$};
 ;
\end{tikzpicture}}
, $M_w \rangle$}  & \text{if } x = \langle M, w \rangle \text{ for a Turing machine $M$ and string $w$}\\\\
\scalebox{.7}{$\langle$ \hspace{-.5cm} \raisebox{-.4cm}{
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt, auto, node distance=2cm, semithick]
      \tikzstyle{every state}=[text=black, fill=none]
      \node[initial,state,accepting] (q0)                    {$q_{\mathrm{acc}}$};
     ;
    \end{tikzpicture}}}
    , ~~~
    \scalebox{.7}{\hspace{-.5cm} \raisebox{-.4cm}{\begin{tikzpicture}[->,>=stealth',shorten >=1pt, auto, node distance=2cm, semithick]
        \tikzstyle{every state}=[text=black, fill=none]
        \node[initial,state] (qrej)                    {$q_{\mathrm{rej}}$};
        \node[state,accepting] (qacc) [right of=qrej]            {$q_{\mathrm{acc}}$};
       ;
      \end{tikzpicture}}$\rangle$ }  & \text{otherwise}.
\end{cases}
\]
Where for each Turing machine $M$, we  define 
\begin{align*}
    M_w = ``&\text{On input } y \\
    &1. \text{   Simulate $M$ on $w$.}\\
    &2. \text{   If it accepts, accept.}\\
    &3. \text{   If it rejects, reject."}
\end{align*}

\item\gradeCorrectFirst $HALT_{\mathrm{TM}} \le_m EQ_{\mathrm{TM}} $ with 
\[
f(x) = \begin{cases}
 \scalebox{.7}{$\langle$ \hspace{-.5cm} \raisebox{-.4cm}{
\begin{tikzpicture}[->,>=stealth',shorten >=1pt, auto, node distance=2cm, semithick]
  \tikzstyle{every state}=[text=black, fill=none]
  \node[initial,state,accepting] (q0)                    {$q_{\mathrm{acc}}$};
 ;
\end{tikzpicture}}
, $M_w \rangle$}  & \text{if } x = \langle M, w \rangle \text{ for a Turing machine $M$ and string $w$}\\\\
\scalebox{.7}{$\langle$ \hspace{-.5cm} \raisebox{-.4cm}{
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt, auto, node distance=2cm, semithick]
      \tikzstyle{every state}=[text=black, fill=none]
      \node[initial,state,accepting] (q0)                    {$q_{\mathrm{acc}}$};
     ;
    \end{tikzpicture}}}
    , ~~~
    \scalebox{.7}{\hspace{-.5cm} \raisebox{-.4cm}{\begin{tikzpicture}[->,>=stealth',shorten >=1pt, auto, node distance=2cm, semithick]
        \tikzstyle{every state}=[text=black, fill=none]
        \node[initial,state] (qrej)                    {$q_{\mathrm{rej}}$};
        \node[state,accepting] (qacc) [right of=qrej]            {$q_{\mathrm{acc}}$};
       ;
      \end{tikzpicture}}$\rangle$ }  & \text{otherwise}.
\end{cases}
\]
Where for each Turing machine $M$, we  define 
\begin{align*}
    M_w = ``&\text{On input } y \\
    &1. \text{   If } y \text{ is not the empty string, reject.}\\
    &2. \text{   Else, simulate $M$ on $w$.}\\
    &3. \text{   If it accepts, accept.}\\
    &4. \text{   If it rejects, reject."}
\end{align*}

\item\gradeCorrect $\{w w \mid w \in \{0,1\}^* \} \leq \emptyset$ and
$f(x) = 1$ for each $x \in \{0,1\}^*$.

\item\gradeCorrect $\emptyset \le_m \{w w \mid w \in \{0,1\}^* \}$ and
$f(x) = 1$ for each $x \in \{0,1\}^*$.


\end{enumerate}

%%%%%%%%%%% PROBLEM 2 %%%%%%%%%%%
\item\textbf{Using mapping reductions} (14 points):
Consider the following computational problems we've discussed
\begin{align*}
A_{TM} &= \{ \langle M, w \rangle \mid M \text{ is a Turing machine, } w \text{ is a string and $M$ accepts $w$}\} \\
HALT_{TM} &= \{ \langle M, w \rangle \mid M \text{ is a Turing machine, } w \text{ is a string and $M$ halts on $w$}\} \\
E_{TM} &=  \{ \langle M \rangle \mid M \text{ is a Turing machine and } L(M) = \emptyset\} \\
EQ_{TM} &= \{ \langle M_1, M_2 \rangle \mid M_1, M_2 \text{ are both Turing machines and } L(M_1) = L(M_2) \}
\end{align*}
and the new computational problem
\[
    Two_{TM} = \{ \langle M \rangle \mid M \text{ is a Turing machine and $M$ accepts exactly one string} \}
\]
NOTE: the intended definition of $Two_{TM}$ is that $M$ accepts exactly two strings. 

\begin{enumerate}
\item[(a)] \gradeCorrect Give an example of a string that is an element of $Two_{TM}$ and a string that is not an element of
$Two_{TM}$ and briefly justify your choices.
\item[(b)] \gradeComplete Prove that $Two_{TM}$ is not decidable by showing that $A_{TM} \leq_m Two_{TM}$.
\item[(c)] \gradeCorrect Give a different proof that $Two_{TM}$ is not decidable by showing that $HALT_{TM} \leq_m Two_{TM}$.
\item[(d)] \gradeComplete Is $Two_{TM}$ recognizable? Justify your answer.
\end{enumerate}

%%%%%%%%%%% PROBLEM 3 %%%%%%%%%%%
\item\textbf{Using mapping reductions} (14 points):
Consider the following computational problems we've discussed
\begin{align*}
A_{TM} &= \{ \langle M, w \rangle \mid M \text{ is a Turing machine, } w \text{ is a string and $M$ accepts $w$}\} \\
HALT_{TM} &= \{ \langle M, w \rangle \mid M \text{ is a Turing machine, } w \text{ is a string and $M$ halts on $w$}\} \\
E_{TM} &=  \{ \langle M \rangle \mid M \text{ is a Turing machine and } L(M) = \emptyset\} \\
EQ_{TM} &= \{ \langle M_1, M_2 \rangle \mid M_1, M_2 \text{ are both Turing machines and } L(M_1) = L(M_2) \}
\end{align*}
and the new computational problem
\begin{align*}
    Evens_{TM} &= \{ \langle M \rangle \mid M \text{ is a Turing machine 
    and any string that $M$ accepts must be even length}\\
    &\text{(but there may even length strings that $M$ doesn't accept)} \}
\end{align*}
\begin{enumerate}
\item[(a)] \gradeCorrect Give an example of a string that is an element of $Evens_{TM}$ and a string that is not an element of
$Evens_{TM}$ and briefly justify your choices.
\item[(b)] \gradeComplete Prove that $Evens_{TM}$ is not decidable by showing that $\overline{HALT_{TM}} \leq_m Evens_{TM}$. {\it This was changed Nov 26.}
\item[(c)] \gradeCorrect Give a different proof that $Evens_{TM}$ is not decidable by showing that $\overline{A_{TM}} \leq_m Evens_{TM}$. {\it This was changed Nov 26.}
\item[(d)] \gradeComplete Is $Evens_{TM}$ recognizable? Justify your answer.
\end{enumerate}


%%%%%%%%%%% PROBLEM 4 %%%%%%%%%%%
\item \textbf{Examples of languages} (7 points):

For each part of the question, use precise mathematical notation or English to define your examples
and then briefly justify why they work.

\begin{enumerate}
    \item\gradeCorrect Two undecidable languages $L_1$ and $L_2$ over the same alphabet
        whose intersection $L_1 \cap L_2$ is decidable, or write {\bf NONE}
        if there is no such example (and explain why).
    
    \item\gradeCorrect A regular language $L_3$ and an unrecognizable
        language $L_4$ over the same alphabet whose set-wise concatenation $L_3 \circ L_4$ is unrecognizable, 
        or write {\bf NONE} if there is no such example (and explain why).
    
    
    \item\gradeComplete A co-recognizable language $L_5$ that is NP-complete,
         or write {\bf NONE} if there is no such example (and explain why).
        Recall the definition: A language $L$ over an  alphabet $\Sigma$ is called {\bf co-recognizable} if its complement,  defined
        as $\Sigma^* \setminus L  = \{ x  \in  \Sigma^* \mid x \notin  L \}$, is Turing-recognizable.
        
\end{enumerate}
    

\end{enumerate}
\end{document}