\documentclass[12pt, oneside]{article}

\input{../resources/discrete-math-packages}

\begin{document}
\begin{flushright}
\StrBefore{\currfilename}{.}
\end{flushright}

\section*{This week's highlights}
\begin{itemize}
\item Evaluate the truth value of a compound proposition given truth values of its constituent variables.
\item Prove propositional equivalences using truth tables
\item Prove propositional equivalences using other known equivalences, e.g.
\begin{itemize}
\item DeMorgan's laws
\item Double negation laws
\item Distributive laws, etc.
\end{itemize}
\item Compute the CNF and DNF of a given compound proposition.
\item Translate sentences from English to propositional logic using appropriate propositional variables and boolean operators.
\item Form the converse, contrapositive, and inverse of a given conditional statement.
\item Decide and justify whether or not a collection of propositions is consistent.
\end{itemize}

\section*{Lecture videos}
Monday: No class in observance of Martin Luther King day.

\href{https://www.youtube.com/watch?v=6KLTpoTpkXo}{A video to reflect on the role of algorithms in
systemic racism. Credit: Safiya Umoja Noble.}

Week 3 Day 1
\href{https://youtube.com/playlist?list=PLML4QilACLk4eQeliamh26adMOwifz8Uj}{YouTube playlist}

Week 3 Day 2
\href{https://youtube.com/playlist?list=PLML4QilACLk7GdezMI_tPsCB2iJ5f7ikp}{YouTube playlist}

\newpage
\section*{Wednesday January 20}

%  DONE
% proposition-def.tex


% logical operators

% logical operators truth tables

% logical operators exercise

% logical equivalence

% logical-equivalence-exercise

% logical-equivalence-example

% truth-table-to-compound-proposition-application

% dnf-cnf definition

% /DONE



\newpage
\section*{Friday January 22}

% hypothesis-conclusion


% logical-equivalence-example-two

% converse-inverse-contrapositive



\newpage

% compound-propositions-translation.tex


\section*{Review quiz questions}
\begin{enumerate}
\item {\bf Wednesday}
% quiz-logic-circuit.tex DONE

% quiz-truth-values-one.tex DONE



\item {\bf Wednesday} 

\newpage
\item {\bf Friday} 

% quiz-truth-values-two.tex

\item {\bf Friday} 

% quiz-consistency.tex


\end{enumerate}
\end{document}
