%! app: TODOapp
%! outcome: TODOoutcome

We will prove that there is no greatest prime number

{\bf Proof} Assume, towards a \underline{BLANK1}, that 
there is a greatest prime number, call it $n_{BIG}$.
In particular, this means that there are finitely many 
primes. Let's label them in order $p_1, \ldots, p_n$ where 
$p_1 = 2$ and $p_n = n_{BIG}$.
Choose $r = $ \underline{BLANK2}. We proved in class
that $r$ is {\bf true}. It remains to show that (under our 
assumption) $r$ is {\bf false}, because
that would complete the contradiction argument.
Define the integer
\[
    C = (p_1 \cdots p_n) + 1
\]
This is a positive integer greater than $1$. 
However, we will show that it does not have any prime
factors and thus is not a product of primes.
By our assumption, the only prime numbers are $p_1, \ldots, 
p_n$. Thus, to show that $C$ does not have any prime
factors means to show that $p_i$ is not a factor 
of $C$ for each value of $i$ from $1$ to $n$. 
Towards a universal generalization, let $i$ be an arbitrary
between $1$ and $n$ (inclusive). We need to prove that $p_i$ is 
not a factor of $C$. By definition of $C$,
\[
    C = p_i ( p_1 \cdots p_{i-1} p_{i+1} \cdots p_n) + 1
\]
so $C \textbf{ div } p_i = p_1 \cdots p_{i-1} p_{i+1} \cdots p_n$
and $C \textbf{ mod } p_i = 1$ (because $p_i > 1$ since it is prime).
Since $C \textbf{ mod } p_i \neq 0$, $p_i$ is not a factor of $C$.
Thus $C$ witnesses that the universal claim is false, 
and we have proved that $r$ is false.




\begin{enumerate}
\item BLANK1
    \begin{enumerate}
        \item universal generalization
        \item proof of existential by witness
        \item direct proof
        \item proof by contrapositive
        \item proof by cases
        \item proof by contradiction
    \end{enumerate}
\item BLANK2
\begin{enumerate}
    \item The least prime number is $2$.
    \item There is a greatest prime number.
    \item There is a least prime number.
    \item Every positive integer greater than $1$ is a product of primes.
    \item Every positive integer has a base expansion.
    \item There is a greatest integer.
    \item There is no greatest integer.
\end{enumerate}
\end{enumerate}