%! app: TODOapp
%! outcome: TODOoutcome

Fill in the blanks in the following proof that, for any equivalence relation $R$ on a set $A$,
\[
\forall a \in A ~([a]_R \neq \emptyset) ~\land ~\forall a \in A ~\exists b \in A~\left( a \in [b]_R \right)
\]

{\bf Proof}: Since this statement is a  \textbf{(a)}$\underline{\phantom{\hspace{1.3in}}}$
we need to prove two goals.

{\bf Goal 1}: we need to show $\forall a \in A ~([a]_R \neq \emptyset) $
{\it Proof of Goal 1}: Towards a \textbf{(b)}$\underline{\phantom{\hspace{1.3in}}}$ 
consider an arbitrary element $a$ in $A$. 
We will work to show that $[a]_R \neq \emptyset$, namely that $\exists x \in [a]_R$.
By definition of equivalence classes, we can rewrite this goal as 
$\exists x \in A ~( ~(a,x) \in R~)$. 
 Towards a \textbf{(c)}$\underline{\phantom{\hspace{1.3in}}}$, consider $x = a$, 
 an element of $A$ by definition. By \textbf{(d)}$\underline{\phantom{\hspace{1.3in}}}$  of $R$, we know that $(a,a) \in R$  and thus the existential quantification has been proved.
 
{\bf Goal 2}: we need to show $\forall a \in A ~\exists b \in A~\left( a \in [b]_R \right)$
{\it Proof of Goal 2}: Towards a \textbf{(e)}$\underline{\phantom{\hspace{1.3in}}}$ 
consider an arbitrary element $a$ in $A$. By definition of equivalence classes, we can rewrite the goal as 
$\exists b \in A ~( ~(b,a) \in R~)$. 
 Towards a \textbf{(f)}$\underline{\phantom{\hspace{1.3in}}}$, consider $b = a$, 
 an element of $A$ by definition. By \textbf{(g)}$\underline{\phantom{\hspace{1.3in}}}$  of $R$, we know that $(a,a) \in R$  and thus the existential quantification has been proved. 
 
 Since both goals have been proved, the statement has been proved. $\square$
 
 
Consider the following expressions as options to fill in the two proofs above. Give your answer as one of the numbers below for each blank a-c. You may use some numbers for more than one blank, but each letter only uses one of the expressions below.

\begin{multicols}{2}
\begin{enumerate}[label=\roman*]
\item universal quantification
\item existential quantification
\item conditional
\item conjunction
\item disjunction
\item exclusive or
\item biconditional
\item exhaustive proof
\item proof by universal generalization
\item proof of existential using a witness
\item proof by cases
\item direct proof
\item proof by contrapositive
\item proof by contradiction
\item reflexivity
\item symmetry
\item transitivity
\end{enumerate}
\end{multicols}
