%! app: TODOapp
%! outcome: TODOoutcome

For each of  the following  system specifications, 
identify the compound propositions  that give their
translations to logic  and then determine if the
translated collection  of compound
propositions is consistent.

\begin{enumerate}
    \item Specification: If the computer is out of memory, then network connectivity is unreliable. No disk errors can occur when the computer is out of memory. Disk
    errors only occur when network connectivity is unreliable.
    
    Translation: $M =$ ``the computer is  out of memory"; $N = $ ``network connectivity
    is unreliable"; $D = $  ``disk errors  can occur".
    
    \begin{enumerate}
        \item \begin{align*} &\neg M \to  N  \\ & \neg D \to M \\ & D \to N \end{align*}
        \item \begin{align*} &M \to  \neg N  \\ & \neg D \wedge M \\ & N \to D \end{align*}
        \item \begin{align*} &M \to  N  \\ &  M \to \neg D \\ & \neg  N \to \neg D \end{align*}
    \end{enumerate}
    
    \newpage
    \item Specification: Whether you think you can, or you think you can't - you're right.
\footnote{Henry Ford}
    
    Translation: $T =$ ``you  think  you  can"; $C = $  ``you  can".
    
    \begin{enumerate}
        \item \begin{align*} &T \to C \\&  \neg T \to \neg C \end{align*}
        \item \begin{align*} &T \wedge C \\  & \neg  T \wedge \neg C \end{align*}
        \item \begin{align*} &T \to \neg T  \\ & C  \to \neg  C \end{align*}
    \end{enumerate}
    
    \item Specification: A secure password must be private and complicated. If
    a password is  complicated then  it will be hard to  remember.  People
    write down hard-to-remember passwords. If a password is written down, it's  not private.   The password is secure.

    Translation: $S =$ ``the password is secure"; $P = $ ``the password is private"; 
    $C = $  ``the password is  complicated"; $H = $ ``the password is hard to remember";
    $W =  $ ``the password is written down".
    
    \begin{enumerate}
        \item \begin{align*} &\neg (P \wedge C) \to \neg  S  \\ & C \to H  \\ & W \wedge H \\ & W \to  \neg P \\ & S \end{align*}
        \item \begin{align*} &(P \wedge  C)  \to S  \\ &  C \to H\\ & W  \to  H \\  & W \to P \\ & S\end{align*}
        \item \begin{align*} & S  \to (P \wedge C)  \\ &  C \to H\\ & H  \to  W \\  & W \to \neg P \\ & S\end{align*}
    \end{enumerate}

\end{enumerate}