%! app: Decidable Languages, Undecidable Languages
%! outcome: Classify language, Find example languages, Define decision problem, Classify decision problem
    

\begin{center}
    \begin{tabular}{|l|l|l|l|}
    \hline
    & Suppose $M$ is  a TM & Suppose $D$ is  a TM & Suppose $E$ is  an
    enumerator  \\
    &that  recognizes $L$  &that  decides $L$  &that enumerates $L$ \\
    \hline
    If string $w$ is in  $L$ then  \ldots  &&& \\
    &&&\\
    &&&\\
    If string $w$ is not in  $L$ then  \ldots  && &\\
    &&&\\
    &&&\\
    \hline
    \end{tabular}
\end{center}


A language $L$ is {\bf recognized by} a Turing machine $M$ means

\vspace{15pt}

A Turing  machine  $M$ {\bf  recognizes} a language $L$ if means

\vspace{15pt}

A Turing machine $M$ is a {\bf decider}  means

\vspace{15pt}

A language  $L$ is {\bf decided by} a Turing  machine  $M$  means

\vspace{15pt}

A  Turing machine $M$ {\bf decides} a language $L$ means

\vspace{15pt}


{\it From Friday's review quiz: }
Which of the following sentences make sense? Which of those are true?

A language is a decider if it always halts.

\vfill

The union of two deciders is a decider.

\vfill

A language is decidable if and only if it is recognizable.

\vfill

There is a Turing machine that isn't decidable.

\vfill

There is a recognizable language that isn't decided by any Turing machine.

\vfill

\newpage


{\bf Claim}: If two languages  (over a fixed alphabet  $\Sigma$) are Turing-recognizable, then  their union  is  as well.

{\bf Proof using Turing machines}:

\vfill

{\bf Proof using nondeterministic Turing machines}: 

\vfill  

{\bf  Proof using enumerators}:

\vfill

\newpage


The first line of a {\bf high-level description} of a Turing machine specifies the input to the machine, which must be a string.
This string may be the encoding of some object or  list of  objects.  

{\bf Notation:} $\langle O \rangle$ is the string that encodes the object $O$.
$\langle O_1, \ldots, O_n \rangle$ is the string that encodes the list of objects $O_1, \ldots, O_n$.

{\bf Assumption}: There are Turing  machines that can be called as subroutines
to decode the string representations of common objects and  interact with these objects as intended
(data structures).
  
For example, since there are algorithms to answer each of the following questions,
by Church-Turing thesis, there is a Turing machine that accepts exactly those strings for which the 
answer to the question is ``yes''
\begin{itemize}
    \item Does a string over $\{0,1\}$ have even length?

    \vfill

    \item Does a string over $\{0,1\}$ encode a string of ASCII characters?\footnote{An introduction to ASCII 
    is available on the w3 tutorial \href{https://www.w3schools.com/charsets/ref_html_ascii.asp}{here}.}

    \vfill

    \item Does a DFA have a specific number of states?

    \vfill

    \item Do two NFAs have any state names in common?

    \vfill

    \item Do two CFGs have the same start variable?

    \vfill

  \end{itemize}

A {\bf computational problem} is decidable iff language encoding its positive problem instances
is decidable.

The computational problem ``Does a specific DFA accept a given string?'' is encoded by the language
\begin{align*}
  &\{ \textrm{representations of DFAs $M$ and strings $w$ such that $w \in L(M)$}\}  \\
  =& \{ \langle M, w \rangle \mid M \textrm{ is a DFA}, w \textrm{ is a string}, w \in L(M) \}
\end{align*}

The computational problem ``Is the language generated by a CFG empty?'' is encoded by the language
\begin{align*}
  &\{ \textrm{representations of CFGs $G$  such that $L(G) = \emptyset$}\}  \\
  =& \{ \langle G \rangle \mid G \textrm{ is a CFG},  L(G) = \emptyset \}
\end{align*}



The computational problem ``Is the given Turing machine a decider?'' is encoded by the language
\begin{align*}
  &\{ \textrm{representations of TMs $M$  such that $M$ halts on every input}\}  \\
  =& \{ \langle M \rangle \mid M \textrm{ is a TM and for each string } w, \textrm{$M$ halts on $w$} \}
\end{align*}


{\it Note: writing down the language encoding a computational problem is only the first step in 
determining if it's recognizable, decidable, or \ldots }
